---
title: "STAT4540_HW4"
author: "John Dennis"
format: pdf
---

```{r,,message=FALSE,echo=FALSE}
library(astsa)
library(tidyverse)
library(tseries)
```

#### Problem 1

Considering the given $AR(2)$, $X_t = -0.9X_{t-2}+W_t$ we can find the roots by grouping and solving for the roots:

$$
\begin{gathered}
X_t = -0.9X_{t-2}+W_t \\
X_t + 0.9X_{t-2} = W_t \\
(1+0.9B^2)X_t = W_t \\ 
\end{gathered}
$$

Once we're here we can solve for the roots by setting $\phi(z)=0$. Note that we replace B with $z \in \mathbb{C}$:

$$
\begin{gathered}
1+0.9z^2 = 0 \\
0.9z^2 = -1\\ 
z^2 = -\frac{10}{9} \\
\implies \boxed{z = \pm \sqrt{\frac{10}{9}}i= \pm1.054i}
\end{gathered}
$$

And because $\pm 1.054$ falls outside the unit circle, this $AR(2)$ process is causal. Now focusing on the ACF:

```{r}
ACF = ts(ARMAacf(ar=c(0,-0.9), lag=48), start=0, freq=12)
tsplot(ACF, type="h", xlab="LAG", ylim=c(-1,1), gg=TRUE, col=4, xaxt="n")
abline(h=0, col=8)
mtext(side=1, at=seq(0,4,by=.5), text=seq(0,48,by=6), cex=.8)
```

#### Problem 2

```{r,echo = FALSE}
df <- read.table("HWARMA.txt", header= TRUE)
```

##### Time Series A

We'll plot the data and then its ACF to look for trends.

```{r,results='hide',echo = FALSE}
tsplot(df$dat.A, ylab = "dat.A")
acf1(df$dat.A)
```

There are no obvious trends and this could potentially be white noise process. We'll fit an $MA(1)$ given the large spike at 1 within the ACF:

```{r,results='hide',echo=FALSE}
model.A <- sarima(df$dat.A, 0,0,1)
```

##### Time Series B

```{r,echo = FALSE}
tsplot(df$dat.B, ylab = "dat.B")
```

##### Time Series C

```{r,echo = FALSE}
tsplot(df$dat.C, ylab = "dat.C")
```

##### Time Series D

```{r,echo = FALSE}
tsplot(df$dat.D, ylab = "dat.D")
```

##### Time Series E

```{r,echo = FALSE}
tsplot(df$dat.E, ylab = "dat.E")
```

##### Time Series F

```{r,echo = FALSE}
tsplot(df$dat.F, ylab = "dat.F")
```
